---
title: "Data Manipulation- Data Wrangling I"
date: September 18th, 2025
output: github_document
---

```{r}
library(tidyverse)
```

```{r}
litters_df =
  read_csv("./data/FAS_litters.csv", na = c("NA", ".", ""))
litters_df =
  janitor::clean_names(litters_df)
```

```{r}
pups_df =
  read_csv("data/FAS_pups.csv", na = c("NA", ".", ""), skip = 3)
pups_df =
  janitor::clean_names(pups_df)
```

# select specific variables from my dataframe

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

# selects a range of variables

```{r}
select(litters_df, group:gd_of_birth)
```

select by removal in df

```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

Let's now rearrange by selecting 

```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())

relocate(litters_df, litter_number)
```

rename by selecting

```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group)
```

what if I only select one variable

```{r}
select(litters_df, group)

pull(litters_df, group)
litters_df$group
```

```{r}
select(pups_df, litter_number, pd_ears, sex)
```

fiter options

```{r}
filter(litters_df, pups_born_alive != 6)
```

Match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Mod8"))
```

Let's filter our missing rows with drop_na() for the entire dataset or for whatever columsn you care about

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```

```{r}
filter(pups_df, sex == 1)
filter(pups_df, sex == 2, pd_walk < 11)
```

## mutating variables (create new + modify exisiting)

```{r}
mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
  )
```

## let's arrange to arrange things

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, desc(group), pups_born_alive)

arrange(litters_df, gd0_weight)
```

# using pipes operators to put things together. Use pipes to avoid nesting codes

```{r}
litters_df =
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(group, litter_number, starts_with("gd")) |> 
  drop_na() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```

```{r}
pups_df =
  read_csv("data/FAS_pups.csv", na = c("NA", ".", ""), skip = 3) |>
  janitor::clean_names() |> 
  filter(sex == 1) |> 
  select(-pd_ears) |> 
  mutate(pd_pivot_gt7 = pd_pivot >= 7)
```


