---
title: "Data Import - Data Wrangling I"
date: September 16th, 2025
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

Let's import a dataset

```{r}
litters_df =
  read_csv("data/FAS_litters.csv")
```

Always look at your data!

```{r}
names(litters_df)
```

Update the names in 'litters_df'

```{r}
litters_df = 
  janitor::clean_names(litters_df)
```

```{r}
litters_df
```

Sometimes skimming data is near?

```{r}
skimr::skim(litters_df)
```

Let's now fix the missing data vlue

```{r}
read_csv("data/FAS_litters.csv", na = c("NA", ".", ""), skip = 5)
```

```{r}
pups_df = 
  read_csv("data/FAS_pups.csv",
           skip = 3,
           na = c("NA", ".", "")
  )
pups_df = 
  janitor::clean_names(pups_df)
```


Alright, what about excel

CSVs are really great but sometimes you get an excel file. Givne the option, always use the csv files.

```{r}
mlb_df =
  read_excel("data/mlb11.xlsx")
```

Import LotR word counts

```{r}
fotr_df =
  read_excel("data/LotR_Words.xlsx", range = "B3:D6")
```

let's now read a SAS file. We'll now import the pulse dataset

```{r}
pulse_df =
  read_sas("data/public_pulse_data.sas7bdat")

pulse_df =
  janitor::clean_names(pulse_df)
```

why do I hate read read.csv so much???

```{r, eval = FALSE}
litters_df_base =
  read.csv("data/FAS_litters.csv")
```

